% Appendix D: Glossary
\chapter{Glossary}
\label{app:glossary}

\begin{description}[style=nextline]

\item[A2A (Agent-to-Agent)]
Protocol for direct communication between AI agents, enabling autonomous agent payments.

\item[ATA (Associated Token Account)]
Solana account derived from a wallet address and token mint, used for holding SPL tokens.

\item[CAIP-2]
Chain Agnostic Improvement Proposal 2, a standard for blockchain network identifiers (e.g., \code{eip155:8453}).

\item[Client]
Any application or agent that requests access to protected resources and submits payments.

\item[EIP-3009]
Ethereum Improvement Proposal enabling gasless token transfers via cryptographic signatures.

\item[EIP-712]
Ethereum standard for typed structured data signing, providing human-readable signing prompts.

\item[ERC-4337]
Account abstraction standard enabling smart contract wallets and gasless transactions.

\item[Facilitator]
Service that handles payment verification and blockchain settlement on behalf of Resource Servers.

\item[HTTP 402]
HTTP status code ``Payment Required,'' used by \tprotocol{} to signal payment requirements.

\item[Jetton]
TON blockchain's fungible token standard, analogous to ERC-20 on Ethereum.

\item[LayerZero]
Cross-chain messaging protocol used by USDT0 for omnichain transfers.

\item[MCP (Model Context Protocol)]
Protocol enabling AI agents to interact with tools and resources, with \tprotocol{} support for paid tools.

\item[Nonce]
Unique value used once to prevent replay attacks; in \tprotocol{}, a 32-byte random value.

\item[OFT (Omnichain Fungible Token)]
LayerZero token standard enabling native cross-chain transfers; USDT0 is an OFT.

\item[PaymentPayload]
JSON structure containing signed payment authorization sent by clients.

\item[PaymentRequired]
JSON structure returned by servers describing acceptable payment methods.

\item[Resource Server]
Service providing protected resources (APIs, content) that require payment for access.

\item[Settlement]
The process of executing a payment transaction on the blockchain.

\item[SPL Token]
Solana Program Library token standard, analogous to ERC-20 on Ethereum.

\item[Stablecoin]
Cryptocurrency designed to maintain stable value, typically pegged to fiat currency (e.g., USDT).

\item[TRC-20]
TRON blockchain's token standard, analogous to ERC-20 on Ethereum.

\item[USDT]
Tether USD, the most widely used stablecoin, available on multiple blockchains.

\item[USDT0]
Tether's omnichain USDT implementation using LayerZero OFT standard.

\item[Verification]
The process of validating a payment signature and parameters before settlement.

\end{description}
